### 2023-06-24 chapter_2/src/manuscript.qmd
oops, forgot ppc's

Are colors important to politics? I argue that they convey information that matters to campaigns and voters. People rely on more than information about policy positions to understand politics [see @pietraszewski_et-al_2015_c;@perez_tavits_2022_pup as examples]. To reduce cognitive load, the affiliation detection system allows for efficient social behavior prediction by associating various cues with a group. In politics, associating Black voters with the Democratic party and White voters with the Republican party is one example [@pietraszewski_et-al_2015_c]. As Americans consistently link the colors red and blue with the Republican and Democratic parties [@elving_2014_npr;@williams_et-al_2022_jomp], as well as their connection to ideology [see @maestre_medero_2022_pr for example], we should explore the degree to which color not only acts as an affiliative cue, but whether it is strong enough to drive individual behavior, and whether it is a strong enough shift to shape electoral strategy.

As a simple yet ubiquitous piece of visual information, color may have more influence over shaping political views than we have given it credit thus far. As visual information is processed automatically and faster than other forms of information [see @ames_et-al_2012_ohsn], in this chapter I argue that it triggers associated neurological pathways and leads us to form a snap-judgment (an attitude without subsequent, more complex, information) that we rely upon for attitudinal and behavioral outcomes. Later chapters will explore the degree to which the inclusion of subsequent substantive policy information overrides this snap-judgement.

To give a preview for this chapter, I propose two studies to explore the claims I make in this chapter. Yard signs are a simple yet effective source of campaign branding for political candidates. They shape the attitudes of those who view them [@makse_et-al_2019_oup], and have enough of an individual-level effect that they shape aggregate attitudes in the form of electoral outcomes [@green_et-al_2016_es]. As yard signs are a simple and static form of campaign advertising, they cannot rely heavily on more complicated forms of information. However, one form of information they rely on and vary in is their color. With these characteristics, yard signs are an ideal candidate both from a statistical standpoint and also substantive. From a statistical standpoint, as they are static forms of party branding they have fewer moving parts, which help isolate the effect that color has on shaping attitudes. From a substantive standpoint, while we understand that yard signs matter to campaigns, we have yet to systematically explore more simple design choices such as color.

In [Study 1](#sec-study-1), I conduct an online experiment that examines whether the public notice color and whether their attitudes are shaped by "Republican Red" and "Democratic Blue" on the yard signs of a fictional candidate. Based on informal conversations with campaign marketing consultants, it appears that campaigns are aware that the colors red and blue betray the partisanship of a candidate to voter. So in [Study 2](#sec-study-2), I examine whether campaigns react to the partisan composition of an electorate as a response to these individual-level patterns. I do this by examining how electoral context shapes variation in using these partisan colors. While these two studies highlight endogenous processes, I am not necessarily arguing that either of these patterns only flow one way. In fact, I believe that there is a necessary feedback loop between the parties and their base of support. Voters support candidates who signal strong allegiance to the party to increase their chances for electoral success [@utych_2020_es] and the parties are motivated to distinguish themselves [@lee_2009;@clifford_2020_pb] as a result of their high levels of ideological and affective polarization [@dietrich_2021_pa;@enders_2021_jop].

From the first study, I find that individuals do shift their perceptions of political candidates in response to the use of the colors red and blue, they consistently and strongly associate Republicans with the color red and Democrats with the color blue, and their willingness to vote for a candidate can be dramatically different depending on whether the color used aligns with the individual's partisan identification.

@@

<!--
A central theme of an edited issue of *The International Journal of Press/Politics* is that visual politics is important and understudied [@lilleker_2019_pm]. Those who are engaged in these questions attribute these challenges to methodological sophistication and the difficult task of interdisciplinary theorizing [@gerodimos_2019_pm; @bucy_joo_2021_ijpp].

The rise of television consumption changed the focus of the medium in the research for political communication scholars [@hall-jamieson_2014_ohpc]. In a similar way, so does the rise of image-based social media and has set new agendas. For several methodological and disciplinary reasons, the visual aspects of television were of little focus in the literature [@bucy_joo_2021_ijpp]. However, the news organizations and politicians are responding to the ubiquitous use of the public to use image-dominant social media platforms like TikTok and Instagram, by joining such platforms. They are relatively active on these platforms as well. Scholars need to make this transition to integrate the role of simple visual information into our theories of political information processing (abbreviated as *pip*) and attitude formation.
-->

@@

What remains unclear is whether these patterns are just simply patterns or whether there are identifiable causes to these patterns. While political psychologists are increasingly interdisciplinary and rely on psychology and neuroscience, it is relatively unclear how color can convey information as complex as politics. Further, it is unclear whether such a simple type of information has enough efficacy to shape political outcomes such as voting intention. This chapter is a first cut at addressing these open questions.

<!--
In western Europe, voters are better at connecting the ideological positioning of a party with the color they use in their branding [@casiraghi_et-al_2022_pp]. The ability to do so is strengthened when the party is longer-surviving and is more prominent [@casiraghi_et-al_2022_pp]. The use of politically-relevant colors activates biases toward pre-existing ideological and partisan preferences among voters in a Spanish sample [@maestre_medero_2022_pr]. And male and female candidates in the United States use different colors -- among other things such as fonts -- in their campaign branding to convey distinguishing information about themselves [@williams_et-al_2022_jomp]. Despite the existing evidence, the particular psychological mechanisms driving this remain unclear. For example, it is unclear how colors may be organized into a cognitive schema that enables voters to quickly access and connect these colors to more complex political attitudes. Furthermore, it is unclear from existing evidence how colors may not just be associated with complex attitudes, but how they may enable political attitude formation and may strengthen existing connections between different political objects. As a jumping-off point, we can use the literature of political information processing (abbreviated as *pip*)\index{political information processing} to establish a common understanding of how political information is processed and how attitudes are formed by the processing of political information.

The first prominent model of *pip* derives from rational choice perspectives. The memory-based model of *pip* views attitudes as a weighted collection of prior information [see @zaller_1992_cup]. As individuals receive new information, they organize it into a schema that is relative to prior objects they already have encoded. By encoding this new information, the model predicts that individuals incorporate this new object with similar objects to form an attitude. This schema then may be more accessible in similar contexts and may then sample from its elements when prompted to express a political attitude. While the Receive-Accept-Sample (RAS) model accepts the view that expressed attitudes are a weighting of information most accessible at the time of attitude expression [@zaller_feldman_1992_ajps], it still presupposes that the weighting is an average of prior information.

The second prominent model challenges this latter point. The online model of *pip* contends that individuals do not evenly weigh information, but that whether they even store it in their long-term memory to access later is biased in the direction of supporting pre-existing attitudes [@lodge_taber_2013_cup]; this phenomenon is referred to as motivated reasoning by the psychology literature [@kunda_1990_pb]. This model suggests that people ignore new information that goes against their prior beliefs and that information confirmatory of their preferences is quicker to access -- referred to as hot cognition [@lodge_taber_2013_cup].

The online model conceptualizes this underlying information encoding and attitude retrieval mechanism as automatic [@lodge_taber_2013_cup]. This occurs due to the information's strong associations with valanced appraisals of the information guiding the attitude [@lodge_taber_2013_cup]. This brings political scientists closer to the dominant conceptualizations among neuroscientists and psychologists concerned with memory retrieval, encoding, and attitude formation [see @fazio_2007_sc]. Namely, that memories are encoded and retrieved quickly based on associations due to their association with valanced appraisals [@kensinger_fields_2022_ohhum]. What remains unexplored is how visual information such as color may prime individuals to engage in motivated reasoning pre-consciously. If they keep their attention on the object, what paths are activated by such information?

-->
## Integrating color into a model of political information processing

@@

I include those questions in the questionnaire due to expectations that they may act as confounds in my hypotheses. As political knowledge intertwines with the strength to which an individual identifies with a political party [@delli-carpini_keeter_1996_yup], I expect that political knowledge is an important confounder in my tests of $H_2$, $H_3$, and $H_4$. Therefore, I include a standard battery (the American National Election Study's battery) for assessing political knowledge.[^color_influence_pre-reg_anon-5] As political knowledge is shaped by levels of interest in politics as well [@delli-carpini_keeter_1996_yup], I include the American National Election Study's question to assess levels of self-reported interest in politics. I additionally include some questions collecting information on participants' ascriptive and descriptive characteristics such as age, education, gender identity, and racial identity, as a number of these are correlates with partisan identification [see @campbell_et-al_1960_jws; @mason_2018_cup].

[^color_influence_pre-reg_anon-5]: A table including the wording of this, and all, measures are included in the Supplementary Information.

@@

There are three more trials that are much like the first trial. What is different between the two other trials is that I vary the amount of color that is on the yard signs (trials 2 and 3). I provide more textual information that deviates from the association of Republicans with red and Democrats with blue (trial 4). Examples of all of these yard signs are included in the appendix. For readers interested in the more technical details, the appendix acommpanying this chapter also includes a table of the data collected in the experiment -- @tbl-study-1-measures; reports full results, posterior predictive checks, and more detailed discussions for all of the models used in the remainder of this section.

@@

Overall, it appears that people viewed the yard signs differently depending on which yard sign they were looking at. Now, I turn to exploring how features about the subjects themselves may shape the way they interact and respond to this visual information on the yard signs.

@@

I examine differences in respondents' reported perceptions of the candidate's partisan affiliation. As participants are asked this question for each treatment they view, the estimand of interest reflects the difference in perceptions of the candidate's partisan affiliation between the yard sign that they saw averaged across respondents. I estimate two ordered logistic regression where the main independent variables of interest are two indicator variables for whether respondents received the "red yard sign" or the "blue yard sign" in each round of Trial 1. The baseline category for both of those indicator variables is if they instead saw the "white yard sign".

@@

I also explore whether these effects may be weaker for those with colorblindness. The results of this model is reported @tbl-colorblindness, which is located in the appendix. The results do not suggest that they are. However, given that I had a very small number of respondents reporting colorblindness, the lack of substantive or statistical effects here are likely because I do not have enough participants with colorblindness in each group to compare.

@@

With the data on the proportion of each of the yard signs using the red and blue colors, I merge this data with the MIT Election Lab Data. This leaves me with data from the House elections between 2016 and 2020. While the data are limited in the number of elections it reflects, it allows me to hold the district composition constant as redistricting occurs after the Census every 10 years. However, each election cycle is different and are often different depending on current events. With these data, I estimate a mixed effect model with partial pooling of the intercepts at the district level and no pooling of the intercepts for the election year level. For less technical readers, this allows me to explore whether the district and within election year differences between choices in yard sign color use. What I am trying to do with this analysis is explore whether or not campaigns in districts that are more partisan will use more red or blue while campaigns in districts that are more moderate will use fewer of those traditionally partisan colors.

@@

>>>>

Now that I have presented some background and have introduced the snap-judgment model of politically-relevant visual information, this chapter is a first attempt to tackle the question at hand. Do colors provide political information that shape attitudes? There are two goals that I am hoping to achieve in this chapter. The first is to convince the reader that the colors red and blue convey affiliative information. Second, that these connections between colors and political groups can shape political outcomes.

This chapter applies the snap-judgment model to a simple case. The use of color on yard signs used by campaigns in elections. I focus on yard signs as they are a common form of campaign marketing and evidence suggests that they can shape individual behavior [@makse_et-al_2019_oup] and electoral outcomes [@green_et-al_2016_es]. Additionally, I start by examining the colors red and blue. While there are other colors that may provide political information -- black and white for Black Lives Matter and Blue Lives Matter movements, the overarching goal of this book is to lay the foundation for a larger research agenda in visual political psychology [see @bucy_joo_2021_ijpp]. Without such foundations, we are studying complex cases and circumstances upon assumptions without a grasp of how, and if they even do, occur in the cases that might seem trivial to some. While there indeed is empirical and experiential evidence suggesting that the Republicans use the color red and Democrats use the color blue [@schloss_palmer_2014_pbr;@williams_et-al_2022_jomp], there is yet to be a theoretical framework that explains how these connections happen at the individual level. It may also be unsatisfying that I do not address *how* these connections came about. While the connections between Republicans with the color red and Democrats with the color blue seemed to quicken after the 2000 Presidential election where news networks covering the election began to consistently use those colors [@elving_2014_npr], it is not clear if this was a deliberate choice or whether this was yet another symptom of an increasingly polarized public who just want easily digestable information about politics so they can move on to other matters in their lives. An undertaking to address such a question is for a different book, perhaps.

To give a preview for this chapter, I propose two studies to explore the claims I make in this chapter. Yard signs are a simple yet effective source of campaign branding for political candidates as they convey information that is often useful for voters -- partisanship [@campbell_et-al_1960_jws]. As yard signs are a simple and static form of campaign advertising, they cannot rely heavily on more complicated forms of information. However, one form of information they rely on and vary in is their color. With these characteristics, yard signs are an ideal candidate both from a statistical standpoint and also substantive for an experimental test of my claims in this chapter. From a statistical standpoint, as they are static forms of party branding they have fewer moving parts, which help isolate the effect that color has on shaping attitudes. From a substantive standpoint, while we understand that yard signs matter to campaigns, we have yet to systematically explore more simple design choices such as color.

In [Study 1](#sec-study-1), I conduct an online experiment that examines whether the public notice color and whether their attitudes are shaped by "Republican Red" and "Democratic Blue" on the yard signs of a fictional candidate. Based on informal conversations with campaign marketing consultants, it appears that campaigns are aware that the colors red and blue betray the partisanship of a candidate to voter. So in [Study 2](#sec-study-2), I examine whether campaigns react to the partisan composition of an electorate as a response to these individual-level patterns. I do this by examining how electoral context shapes variation in using these partisan colors. While these two studies highlight endogenous processes, I am not necessarily arguing that either of these patterns only flow one way. In fact, I believe that there is a necessary feedback loop between the parties and their base of support. Voters support candidates who signal strong allegiance to the party to increase their chances for electoral success [@utych_2020_es] and the parties are motivated to distinguish themselves [@lee_2009;@clifford_2020_pb] as a result of their high levels of ideological and affective polarization [@dietrich_2021_pa;@enders_2021_jop]. Again, to tease out the direction of blame is more appropriate for another time.

In the first study, I find that individuals do shift their perceptions of political candidates in response to the use of the colors red and blue, they consistently and strongly associate Republicans with the color red and Democrats with the color blue, and their willingness to vote for a candidate can be dramatically different depending on whether the color used aligns with the individual's partisan identification. The second study suggests that candidates in districts where Democrats have a higher vote share use more blue on their yard signs while also using less red. The implications of this are that the colors red and blue convey political information. More generally, I suspect other colors and more complex forms of visual information such as symbols are capable of conveying political information as well. Perhaps most interesting among these results is that the color red and blue alone can make individuals willing to vote for a candidate. The association between the colors red and blue are strong enough that individuals express a willingness to cast a vote for a candidate based on the color of their yard sign without confirming any of their policy stances or their partisan affiliation. These tendencies, even if based on a hunch for the folks on the ground, seem to lead to descriptive differences in the use of these colors as it guides strategy. The following chapters dig a bit deeper into the idea that colors can shape attitudes by pitting the information individuals get from color against information gleaned from explicit policy stances.

@@

A central theme of an edited issue of *The International Journal of Press/Politics* is that visual politics is important and understudied [@lilleker_2019_pm]. Those who are engaged in these questions attribute these challenges to methodological sophistication and the difficult task of interdisciplinary theorizing [@gerodimos_2019_pm; @bucy_joo_2021_ijpp].

The rise of television consumption changed the focus of the medium in the research for political communication scholars [@hall-jamieson_2014_ohpc]. In a similar way, so does the rise of image-based social media and has set new agendas. For several methodological and disciplinary reasons, the visual aspects of television were of little focus in the literature [@bucy_joo_2021_ijpp]. However, the news organizations and politicians are responding to the ubiquitous use of the public to use image-dominant social media platforms like TikTok and Instagram, by joining such platforms. They are relatively active on these platforms as well. Scholars need to make this transition to integrate the role of simple visual information into our theories of political information processing (abbreviated as *pip*) and attitude formation.

@@

In western Europe, voters are better at connecting the ideological positioning of a party with the color they use in their branding [@casiraghi_et-al_2022_pp]. The ability to do so is strengthened when the party is longer-surviving and is more prominent [@casiraghi_et-al_2022_pp]. The use of politically-relevant colors activates biases toward pre-existing ideological and partisan preferences among voters in a Spanish sample [@maestre_medero_2022_pr]. And male and female candidates in the United States use different colors -- among other things such as fonts -- in their campaign branding to convey distinguishing information about themselves [@williams_et-al_2022_jomp]. What remains unclear is whether these patterns in the United States are just simply patterns or whether there are identifiable causes to these patterns. While political psychologists are increasingly interdisciplinary and rely on psychology and neuroscience, it is relatively unclear how color can convey information as complex as politics. Further, it is unclear whether such a simple type of information has enough efficacy to shape political outcomes such as voting intention. This chapter is a first cut at addressing these open questions.

## Integrating color into a model of political information processing

@@

I include those questions in the questionnaire due to expectations that they may act as confounds in my hypotheses. I include some questions collecting information on participants' ascriptive and descriptive characteristics such as age, education, gender identity, and racial identity, as a number of these are correlates with partisan identification [see @campbell_et-al_1960_jws; @mason_2018_cup]. Though political knowledge does predict the strength of an individual's partisan identification [@lodge_taber_2013_cup], attention paid to politics is likely a more important confounding variable here. The purpose of color in politics is to act as a heuristic for people (a cognitive shortcut). If individuals are aware of the tendency for Republicans to use and be associated with the color red and Democrats with the color blue, the degree to which they know factual information about politics is not entirely relevant. On the other hand, those that pay no attention to politics, whatsoever, may be less likely to be aware of such associations. Additionally, those who express less interest in politics are also likely to identify as more moderate when asked about their ideology and partisan identity [@klar_krupnikov_2016_cup].

@@

There are three more trials that are much like the first trial. What is different between the two other trials is that I vary the amount of color that is on the yard signs (trials 2 and 3). I provide more textual information that deviates from the association of Republicans with red and Democrats with blue (trial 4). Examples of all of these yard signs are included in the appendix. For readers interested in the more technical details, the appendix acommpanying this chapter also includes a table of the data collected in the experiment -- @tbl-study-1-measures; reports full results, posterior predictive checks, and more detailed discussions for all of the models used in the remainder of this section.^[Unless noted otherwise, the reflection of uncertainty in the models are credible intervals at the 90% level. This should not be equated with the frequentist practice of often reporting confidence intervals at the 95% level as credible intervals at this level often are extremely unstable and, more importantly, are interpreted differently. For example, a credible interval at the 90% level can be interpreted as: the probability that the true effect is in that interval. This contrasts with a confidence interval used in the frequentist framework which is interpreted as: upon repeated samples, 90% of the confidence intervals contain the true effect. As I am using two-sided credible intervals here, at the 90% level, the probability that 0 would be the true value given the data would be less than 5% as it would be on one side of my posterior distribution.]

@@

Overall, it appears that people viewed the yard signs differently depending on which yard sign they were looking at. Now, I turn to exploring how features about the subjects themselves may shape the way they interact and respond to this visual information on the yard signs. Though, I will note that the data on the timing and cursor location is much more noisy than I had originally anticipated. However, as the tests of the next few questions demonstrate, there seem to be systematic differences between individuals that saw either a red, white, or blue yard sign despite holding everything else but the color constant between them. While I am a bit skeptical of the data here as it is noisy, I think the next peices of empirical evidence suggest support for this hypothesis as well.

@@

I examine differences in respondents' reported perceptions of the candidate's partisan affiliation. Evidence supporting $H_2$ would appear as differences between those who saw the white yard signs and the "partisan color" yard signs in reported perceptions of the partisanship of the candidates on average. I estimate two ordered logistic regression where the main independent variables of interest are two indicator variables for whether respondents received the "red yard sign" or the "blue yard sign" in each round of Trial 1. The baseline category for both of those indicator variables is if they instead saw the "white yard sign".

@@

I also explore whether these effects may be weaker for those with colorblindness. The results of this model is reported @tbl-colorblindness, which is located in the appendix. The results do not suggest that they are. However, given that I had a very small number of respondents reporting colorblindness in my sample, the lack of substantive or statistical effects here are likely because I do not have enough participants with colorblindness in each group to compare.

@@

With the data on the proportion of each of the yard signs using the red and blue colors, I merge this data with the MIT Election Lab Data. This leaves me with data from the House elections between 2018 and 2022. While the data are limited in the number of elections it reflects, it allows me to hold the district composition constant as redistricting occurs after the Census every 10 years. However, each district and state is different in their laws surrounding elections and are different in terms of political context. I want to examine whether there are meaningful differences between districts that are partisan "strongholds" (Democrats or Republicans tend to get elected in a district) relative to districts that are much less so. To do this, I include a model with intercepts with no pooling. That is a model that includes an intercept for each state-district unit. Though my outcomes of interest (proportion of the yard sign that is red or blue) are continous (the outcomes can take any real number, so not an integer), the outcomes are bounded between 0 and 1. While the traditional approach may be to use ordinary least squares, such a technique often can lead to poor model fit and introduce systematic error into my model and lead to results that are quite unstable. To not bog down readers with a lot of technical jargon, the model that I end up using is called a ordered beta regression which allows me to efficiently and accurately predict outcomes that can be 0 or 1 or any real number between 0 and 1 [@kubinec_2022_pa]. For readers in a more technical discussion about why I chose this over other techniques in the beta regression family, I encourage you to look at the appendix.

Once I have fit these models, I use the `marginaleffects` [@marginaleffects] package to calculate the grand mean of the predicted proportion of the use of the color red or blue for these yard signs as explained by the 5-year rolling average Democratic candidates' vote share. The grand mean here just reflects that I am not showing the predicted proportion of the use of these outcomes for each state-district, but rather I am just averaging across all of those to provide a single line. These results are included in @fig-study-2-pred.

@@

The results demonstrate that districts where Democrats are historically more successful see yard signs with a lower proportion of red on them, while those same districts see more Blue on them. Again, these results do not claim that campaigns are doing this either because voters are encouraging them to take this strategy or that they are doing it as a way to proactively get voters to do it. To tease out which of those questions is the right one requires more sophisticated approaches and data that could allow me to leverage redistricting to see how these outcomes change due to some unrelated, external shock. The data available to me do not allow me to do that at this time. Furthermore, teasing out such an association is not a primary goal of this chapter.

## Conclusions

This chapter argues that the colors red and blue provide useful information to voters about politics. Specifically, the colors red and blue provide information about a candidates' partisanship. Second, the chapter argues that the information gleaned from these colors can shape how people engage with politics. While I had a number of other sets of expectations from this chapter, these were the two primary questions I wanted to address in the chapter.

Evidence from my experiment suggests that the use of the colors red and blue convey partisan information, while a third color like white provides very little useful political information. The experiment provides evidence that individuals see candidates that use red on a yard sign as more Republican and candidates that use blue as more Democratic. Evidence from the experiment also suggested that using less of these partisan colors and using both partisan colors do make it a bit harder for individuals to make those same associations. Further, the experiment demonstrated that these associations that we make with color and partisanship are powerful enough that they can shape whether people express a willingness to vote for a candidate. That is, individuals that identify with the Republican party are much more likely to vote for a candidate with a red yard sign than a white yard sign and Democrats are much more likely to vote for a candidate with a blue yard sign than yard signs with other colors.

Since it appears that colors can influence vote choice, I turn to data that capture the real world messiness to examine whether or not campaigns are responsive to such patterns. It appears that they are. The empirical evidence suggest that campaigns use more blue on their yard signs if it is a district that has a higher vote share for Democrats the five previous years and use red much less often in those same circumstances. This fits with strategies communicated to me by a campaign marketing firm.

What this chapter demonstrates is that the colors red and blue convey information and they shape outcomes. It appears as though that it does not simply shape outcomes for individuals such as their personal attitudes or their vote choice in a lab, but that it has enough of an aggregative effect that can shape strategy for campaigns. The next chapter seeks to dig into this latter claim a bit more. The next chapter examines whether the snap-judgment model holds up in more complex settings. Specifically, the next chapter examines whether color can shape whether and how we have conversations about politics with others and the degree to which color can be a tool inhibiting political persuasion.
